---
title: "Bollt_FinalProject_Thermocline"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The crux of this project is to compare how the thermoclines on each of the 9 Wisconsin lakes in the NTL-LTER dataset have moved over the course of the 35 years of data. I am interested in this question because I have an interest in coldwater fisheries that rely on thermoclines to survive the summer weather. Perhaps climate change is affecting where the thermocline sets up, and therefore shrinking the available summer habitat of trout.
```{r}
#Setting up session and loading all packages I think I might need
getwd()
library(tidyverse)
library(lubridate)
library(ggplot2)
library(multcompView)
library(nlme)
library(lsmeans)

#Reading in the raw lake dataset
NTLR_raw <- read.csv("V:/ENV_872_Project_Directory/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
View(NTLR_raw)
```
Next steps (document everything with comments as you work):
Change the date column to dates
Filter to generate a dataset for each of the nine lakes.
Determine how to set a thermocline benchmark (July average?)
Determine what data to compare to over time (see Kateri's comment)
Run that comparison for each lake
Figure out how to compare each of the lakes to eachother to look for overall trends
Make a bunch of graphs
Write the report
Go fishing