---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("./Data/Processed/NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d") # capital Y means a four-digit year
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
View(PeterPaul.chem.nutrients)
```

## GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

*****Ingredients for making a ggplot graph:********
ggplot (dataframe, aes(assigning x and y variables), color, etc.)) + 
  geom_x( aes(if you want))

### Plotting continuous variables over time: Scatterplot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))  # another way to make the same graph; this one puts it in environments
print(O3plot) #lets you knit a dataframe into a markdown document

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue") # color goes outside of aesthetics
print(O3plot2)

# Add additional variables
PMplot <- 
  ggplot(EPAair, aes(x = month, y = PM25, shape = as.factor(year), color = Site.Name)) + 
  geom_point()
  # graphs months vs pm25; changes shape based on year, you need to treat year as a factor; sets color to site.name
print(PMplot)

# Separate plot with facets so that different years are separate graphs, and not separate shapes for each year's points
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y = PM25, shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3)  #breaks the graph into different graphs for each site name; puts the three graphs into different rows
print(PMplot.faceted)

PMplot.faceted2 <-
  ggplot(EPAair, aes(x = month, y = PM25)) +
  geom_point() +
  facet_grid(Site.Name ~ year)   # ~ means "by"; puts both 2017 and 2018 in the same timeseries
print(PMplot.faceted2)

# Filter dataset within plot building
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) +  # just making a graph of Blackstone w subset
  geom_point() +
  geom_line()   #connects the dots between points
print(O3plot.Blackstone)

# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
# Plot surface temperatures by day of  year. 
# Color your points by year, and facet by lake in two rows.
peterpaul.nutrient.plot <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = daynum, y = temperature_C, color = year4)) +
  geom_point() +
  facet_wrap(vars(lakename), nrow = 2)
print(peterpaul.nutrient.plot)
  
  

#2. 
# Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5
ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y = temperature_C, color = depth)) +
  geom_point(size= 0.5)

```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +        # zooms into the area where our nonoutlier data is
  ylim(0, 20)
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
tempvsdepth <- 
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()  #switches x and y axes 
print(tempvsdepth)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() #+
  geom_smooth(method = lm) +     # draws a best fit line 
  geom_abline(aes(slope = 16, intercept = 0), lty = 2) # draws a line with a slope of 16 and an intercept of 0 and a dashed line
print(NvsP)

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by site. 
# Add a line of best fit for the linear regression of these variables.
lesson9_plot <- ggplot(EPAair, aes(x = PM2.5, y = Ozone, color = Site.Name)) +  # the color is defined inside the aesthetics,each site gets its own trend line because aesthetics were defined at the top
  geom_point() +
  geom_smooth(method = lm)
print(lesson9_plot)

  
```

### Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>%    # grouping by multiple factors at once
  summarise(sd = sd(concentration), 
            mean = mean(concentration))

Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = as.factor(nutrient)), # We use fill because we want to fill the bars
           position = "dodge", stat = "summary", fun.y = "mean")             # position = dodge spreads out the bars, stat = summary in combo with fun.y = mean tells geom_bar to compute and then graph mean on the y axis
print(Nutrientplot)

Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = as.factor(nutrient))) + # graph of lakename vs mean colored by nutrient
  geom_bar(stat = "identity", position = "dodge") + # what does the stat command do?  The stat function is telling geom_bar what to graph on the y axis
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), # how do we specify error bars? By explictly specifying them in aesthetics
                 position = "dodge")
print(Nutrientplot2)

# Are there more effective ways to produce summary stats for categories?

# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
print(Nutrientplot3)

# Dot plot # plots equal values next to eachother
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") #
print(Nutrientplot4)

# Violin plot # a line instead of dots, but similar to a dot plot
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") +   # specify colors or all the lines will be black
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") 
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)   # same plot as above

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
# Create separate bars for each site.
EPAair.summary <-

ggplot(EPAair)+
  geom_bar(aes(x = Site.Name, y = PM25, fill= as.factor(year)), position = "dodge", stat = "summary", fun.y = "mean") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd))

# this code isn't generating error bars because you didn't define what the mean and standard deviation are
# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.

```
